% ============================================================
% HEAS — WSC 2026 Submission
% Track: Modeling Methodology
%
% INSTRUCTIONS FOR COMPILATION:
%   1. Download the WSC 2026 Authors Kit from:
%      https://meetings.informs.org/wordpress/wsc2026/authorsubmission/
%   2. Copy wsc.cls (and any support files) into this directory.
%   3. Run: pdflatex paper && bibtex paper && pdflatex paper && pdflatex paper
%
% REQUIREMENT CHECKLIST:
%   [x] Abstract <= 150 words, single paragraph, no keywords, no math symbols
%   [x] Title bold ALL CAPS (handled by wsc class)
%   [x] Section titles numbered, bold, ALL CAPS
%   [x] Times New Roman 11pt (handled by wsc class)
%   [x] 5--12 pages
%   [x] Author biographies mandatory (one paragraph + email each)
%   [x] No footnotes
%   [x] No page numbers (handled by wsc class)
%   [x] U.S. English throughout
%   [ ] Verify page count after final compile
% ============================================================

\documentclass{wsc}  % obtain from WSC 2026 Authors Kit

\usepackage{booktabs}      % professional table rules
\usepackage{listings}      % code listings
\usepackage{microtype}     % better text justification
\usepackage{url}           % URL formatting
\usepackage{amsmath}       % math (if needed)
\usepackage{graphicx}      % figures
\usepackage{pifont}        % \cmark / \xmark for Table 7
\usepackage{multirow}      % table cell merging
\usepackage[hidelinks]{hyperref}  % PDF links without boxes

% ---- Code listing style (8pt monospace, no line numbers) ---
\lstset{
  basicstyle=\ttfamily\small,
  columns=flexible,
  keepspaces=true,
  frame=single,
  framerule=0.4pt,
  xleftmargin=6pt,
  xrightmargin=6pt,
  breaklines=true,
  showstringspaces=false,
  numbers=none
}

% ---- Convenience symbols -----------------------------------
\newcommand{\cmark}{\ding{51}}   % check mark
\newcommand{\xmark}{\ding{55}}   % cross mark

% ============================================================
\begin{document}

\title{HEAS: A HIERARCHICAL EVOLUTIONARY AGENT-BASED SIMULATION FRAMEWORK
       FOR MULTI-OBJECTIVE POLICY SEARCH}

% Replace with actual author names and affiliations before submission.
% WSC uses the format shown in wsc.cls; adjust as needed.
\author{
  \authorname{[First Author]}\\
  \authoraddr{[Department, Institution]}\\
  \authoraddr{[City, State, Country]}\\
  \authoremail{[email@institution.edu]}
  \and
  \authorname{[Second Author]}\\
  \authoraddr{[Department, Institution]}\\
  \authoraddr{[City, State, Country]}\\
  \authoremail{[email@institution.edu]}
}

\maketitle

% ============================================================
\begin{abstract}
Agent-based frameworks provide no native support for the pipeline that policy
research demands: multi-objective evolutionary search, multi-scenario tournament
comparison, and statistically rigorous multi-run validation. Researchers bridge
this gap with bespoke coupling code---100--200 lines per project---that is
non-reusable and prone to silent metric divergence. HEAS (Hierarchical
Evolutionary Agent-based Simulation) eliminates this overhead through four
architectural commitments: a Layer/Stream/Arena composition model; a uniform
metric contract consumed identically by the evolutionary algorithm, tournament
scorer, and bootstrap CI pipeline; native NSGA-II integration with Pareto front
output and hypervolume tracking; and a Tournament primitive with four voting
rules, sample complexity certification, and noise stability analysis. Against a
competent Mesa~3.3.1 reimplementation of an ecological model, HEAS requires 5
lines of coupling code versus 160 (97\% reduction). Two case studies validated
across 30 independent runs confirm framework correctness and reproducibility.
\end{abstract}

% ============================================================
\section{INTRODUCTION}

160 lines of coupling code---that is the overhead a competent Mesa programmer
writes to connect a predator-prey simulation to NSGA-II, tournament comparison,
and bootstrap confidence intervals. HEAS reduces this to five lines. The rest
of this paper explains why the gap is structural, not stylistic.

Agent-based models (ABMs) have become a primary tool for studying emergent
behavior in complex adaptive systems---from ecological communities and financial
markets to public health interventions and regulatory policy design. Using an
ABM for \emph{policy search}---finding the configuration of agent parameters or
institutional rules that optimizes a multi-dimensional objective---requires
infrastructure that no general-purpose ABM framework currently provides.

\subsection{The Coupling Code Problem}

Consider a researcher using Mesa to study regulatory policy in a multi-firm
economy. She wants to (1)~evolve a Pareto-optimal tax regime using NSGA-II,
(2)~compare the evolved regime against a reference policy across 32 independent
economic scenarios using tournament voting, and (3)~report 30-run bootstrap
confidence intervals to establish statistical credibility. Each step reads the
same welfare metric from the same model, yet in Mesa each step is an independent
code path:

\begin{itemize}
  \item The \textbf{DEAP fitness function} extracts welfare from the model's
        \texttt{DataCollector} after each run.
  \item The \textbf{tournament scorer} extracts welfare from a separate
        episode-evaluation function---possibly with different aggregation
        (mean vs.\ final value).
  \item The \textbf{bootstrap CI loop} extracts welfare from a third data
        collection loop.
\end{itemize}

When the researcher adds a second objective---welfare inequality---she must edit
the DataCollector reporter, the DEAP fitness function, the tournament scorer, and
the CI loop, keeping all four in sync manually. The framework provides no
enforcement. A silent divergence---the EA optimizing mean welfare while the
tournament evaluates final-step welfare---can persist undetected for weeks and
invalidate a publication.

This pattern, which we call \emph{coupling code}, appears in every ABM-based
policy search project we are aware of~\cite{pangallo2019best,capri2023transport}.
It is not a failure of Mesa or NetLogo---these frameworks were designed for
exploratory, single-run agent modeling. But as ABMs are increasingly deployed
for optimization and evaluation tasks, the coupling code burden has become a
significant source of research friction and reproducibility failures.

\subsection{What Existing Frameworks Provide}

Mesa~3.x~\cite{kazil2020mesa} provides \texttt{DataCollector} for per-step
logging, \texttt{batch\_run} for parameter sweeps, and \texttt{SolaraViz} for
visualization. It has no native support for multi-objective EA, Pareto front
output, tournament voting, or bootstrap CI.
NetLogo's~\cite{wilensky1999netlogo} BehaviorSpace supports discrete grid sweeps,
not gradient-free optimization over continuous gene spaces.
Repast4Py~\cite{collier2022repast4py} targets HPC distributed ABM but provides
no EA integration or tournament evaluation.
ABIDES~\cite{byrd2020abides} is domain-specific to financial markets.
External optimization libraries (PyGMO~\cite{biscani2020pagmo},
Optuna~\cite{akiba2019optuna}) accept Python callables but require the researcher
to write the coupling for each new project.

\subsection{Contributions}

We present \textbf{HEAS} (Hierarchical Evolutionary Agent-based Simulation),
a Python framework that addresses the coupling code problem through four
architectural commitments:

\begin{description}
  \item[C1---Layered Composition.] The \texttt{Layer}/\texttt{Stream}/\texttt{Arena}
    triad formalizes multi-level hierarchy as a first-class abstraction. Layers
    declare their own state and step logic; Streams declare how they couple.
    Hierarchy is visible to the framework, enabling independent testing and
    structural reuse across studies.

  \item[C2---Uniform Metric Contract.] Every \texttt{Layer} implements
    \texttt{metrics\_episode()}, returning a \texttt{dict[str, float]} of
    namespaced keys (e.g., \texttt{agg.mean\_biomass}). This dictionary is the
    \emph{single source of truth} consumed identically by the EA fitness
    function, tournament scorer, and statistical CI pipeline. Adding a metric is
    a one-line change in one file, propagating to all consumers with no
    additional synchronization.

  \item[C3---Native Multi-Objective EA.] \texttt{run\_optimization\_simple()}
    wraps DEAP's~\cite{fortin2012deap} NSGA-II~\cite{deb2002nsga2} with
    per-run seed management, parallel episode evaluation, Pareto front output,
    and hypervolume tracking. The researcher writes only the objective function;
    no DEAP knowledge is required.

  \item[C4---Tournament Evaluation.] The \texttt{Tournament} class implements
    four voting rules (argmax, majority, Borda, Copeland), sample complexity
    certification, and noise stability analysis as framework-level primitives.
    Multi-scenario policy comparison becomes a reproducible, statistically
    certified operation rather than a bespoke script.
\end{description}

Section~5 reports a direct LOC comparison confirming 97\% coupling code
reduction versus a competent Mesa~3.3.1 implementation.

\subsection{Paper Organization}

Section~2 surveys related work. Section~3 describes the HEAS architecture.
Section~4 presents two case studies demonstrating composability across
structurally different domains. Section~5 reports evaluation. Section~6
concludes.

% ============================================================
\section{RELATED WORK}

\subsection{Agent-Based Simulation Frameworks}

\textbf{Mesa}~\cite{kazil2020mesa} is the dominant Python ABM framework,
providing grid and network spaces, agent scheduling, \texttt{DataCollector} for
per-step metric logging, and \texttt{SolaraViz} for browser-based visualization.
Mesa is the right tool for exploratory, spatially-oriented single-run modeling.
Its \texttt{DataCollector} is a pull-based logger: reporters are declared at
initialization and queried at each step, producing a time-series DataFrame.
This design couples metric collection to the model's internal state; each
downstream consumer must write its own extraction function with no framework
mechanism to verify consistency. HEAS complements rather than replaces
Mesa---for spatial ABMs with rich agent interactions and visualization needs,
Mesa remains the preferred choice.

\textbf{NetLogo}~\cite{wilensky1999netlogo} provides BehaviorSpace for
parameter sweeps over a discrete grid. BehaviorSpace does not support continuous
gradient-free optimization, produces no Pareto front, and cannot be used for
multi-scenario tournament comparison. Its Java runtime is not directly composable
with Python statistical pipelines.

\textbf{Repast4Py}~\cite{collier2022repast4py} targets high-performance
distributed ABM on HPC systems. Repast's contribution is scalability; HEAS's
contribution is composability---they address orthogonal concerns.

\textbf{ABIDES}~\cite{byrd2020abides} provides a financial market simulation
framework with message-passing agents calibrated to real exchange data. It is
domain-specific and not designed for policy parameter optimization or
multi-scenario comparison.

\subsection{Optimization over Simulations}

\textbf{PyGMO/Pagmo}~\cite{biscani2020pagmo} provides high-performance
multi-objective optimization accepting any Python callable as the objective.
Integration with an ABM must be written by hand per project---reproducing the
coupling code that HEAS eliminates. \textbf{Optuna}~\cite{akiba2019optuna}
supports simulation objectives but is designed for single-objective Bayesian
search, not multi-objective Pareto search with tournament evaluation.
\textbf{DEAP}~\cite{fortin2012deap}, which HEAS uses internally, provides
NSGA-II but requires the user to wire all infrastructure. HEAS wraps DEAP with
per-run seeding, parallel evaluation, and hypervolume tracking, exposing a
single-call interface that requires no DEAP knowledge.

\subsection{Policy Search in Agent-Based Models}

Domain-specific works have combined ABMs with evolutionary search.
Pangallo et al.~\cite{pangallo2019best} apply genetic algorithms to a
macroeconomic ABM using custom fitness extraction---a representative example
of the coupling code pattern HEAS eliminates.
Zheng et al.~\cite{zheng2022aieconomist} learn tax policies in the AI
Economist using deep RL; this requires a differentiable simulation. HEAS
targets the complementary setting: stochastic, non-differentiable, episodic
simulation where gradient-free Pareto search is appropriate.
Capr\`{i} et al.~\cite{capri2023transport} evaluate transportation policies
using NetLogo with external Python scripts, manually maintaining metric
consistency---the exact divergence risk that HEAS's metric contract prevents.

\subsection{Gap Summary}

Table~\ref{tab:related} summarizes the capability gap that HEAS addresses.
HEAS occupies the intersection of these lines of work: an ABM runtime with
layered hierarchy, connected to evolutionary search via a uniform metric
contract, augmented with tournament evaluation and statistical CI as
framework-level primitives---a combination not available in any existing tool.

\begin{table}[t]
\caption{Framework capability comparison.
  \cmark~=~native; \xmark~=~not provided; ---~=~not applicable.}
\label{tab:related}
\centering\small
\begin{tabular}{lccccc}
\toprule
Framework      & Hier. & Metric & EA  & Tourn. & CI \\
\midrule
Mesa~3.x       & \xmark & \xmark & \xmark & \xmark & \xmark \\
NetLogo        & \xmark & \xmark & \xmark & \xmark & \xmark \\
Repast4Py      & \xmark & \xmark & \xmark & \xmark & \xmark \\
ABIDES         & \xmark & \xmark & \xmark & \xmark & \xmark \\
PyGMO + Mesa   & ---    & \xmark & \cmark & \xmark & \xmark \\
\textbf{HEAS}  & \cmark & \cmark & \cmark & \cmark & \cmark \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{THE HEAS FRAMEWORK}

HEAS is implemented in Python~3.11 and depends on DEAP for evolutionary
algorithms, NumPy for numerical computation, and SciPy for statistical
utilities. All components---EA algorithm, voting rule, statistical
estimator---can be replaced without changing simulation or metric contract code.

\subsection{Layered Composition}

The fundamental abstraction is the \texttt{Layer}, a unit of simulation logic
with a four-method contract:

\begin{lstlisting}[language=Python]
class Layer:
    def step(self) -> None:
        """Advance state by one discrete timestep."""
    def metrics_step(self) -> dict[str, float]:
        """Per-timestep metrics: fed to dashboards."""
    def metrics_episode(self) -> dict[str, float]:
        """End-of-episode aggregates: fed to EA,
           tournament, and CI pipeline."""
    def reset(self) -> None:
        """Reset to initial conditions."""
\end{lstlisting}

The contract separates \emph{what a layer computes} from \emph{how its outputs
are used}. Any downstream component reads the same \texttt{metrics\_episode()}
dictionary using the same key string. This is the metric contract: a single
source of truth enforced by the interface, not by convention.

Layers compose into \textbf{Streams}---named groups of layers forming a
coherent sub-simulation---and Streams compose into an \textbf{Arena}, which
orchestrates step ordering, inter-stream signal routing, and episode-level
aggregation. The naming convention \texttt{stream\_name.metric\_name}
(e.g., \texttt{agg.mean\_biomass}) ensures globally unambiguous keys regardless
of Arena size.

\subsection{Why Hierarchy Matters}

Existing frameworks represent agents as flat objects sharing a global model
state. When a researcher adds climate forcing, spatial fragmentation, and an
aggregation layer to a predator-prey model, these appear as attributes of the
Mesa \texttt{Model} class---invisible to the framework and impossible to reuse,
test in isolation, or compose into a new study.

In HEAS, the same model is a five-stream Arena:

\begin{lstlisting}
EcoArena
 +-- ClimateStream   (temperature, shock events)
 +-- LandscapeStream (patch quality, fragmentation)
 +-- PreyStream      (logistic growth, dispersal)
 +-- PredStream      (predation, mortality)
 +-- AggStream
      -> agg.mean_biomass, agg.cv, agg.extinct
\end{lstlisting}

Each stream can be developed, validated, and replaced independently. Swapping
the climate model requires only modifying \texttt{ClimateStream.step()}---the
EA, tournament, and AggStream see no change because they interact only through
the metric contract.

\subsection{Native Multi-Objective EA Integration}

\texttt{run\_optimization\_simple()} connects DEAP's NSGA-II to any objective
defined via the metric contract:

\begin{lstlisting}[language=Python]
result = run_optimization_simple(
    objective_fn=eco.trait_objective,
    n_genes=2,
    gene_bounds=[(0.0, 1.0), (0.0, 1.0)],
    pop_size=20, n_generations=10, seed=42,
)
# result["hof_fitness"] -- Pareto front
# result["hv"]          -- hypervolume
# result["log"]         -- per-generation stats
\end{lstlisting}

The wrapper handles DEAP's creator/toolbox registration, crossover and mutation
wiring, hall-of-fame management, and hypervolume computation. A researcher
adds NSGA-II to a new simulation by writing only the objective function
(typically 5--10 lines reading from \texttt{run\_many()} results) and calling
\texttt{run\_optimization\_simple()}. The EA algorithm is itself replaceable:
\texttt{run\_ea()} also accepts CMA-ES and single-objective hillclimbing.
Section~\ref{sec:ablation} shows that simple hillclimbing outperforms NSGA-II
on a 2-gene landscape---confirming that algorithm-agnosticism is not only an
architectural claim but a practical benefit.

\texttt{run\_many()} executes $n$ episodes in parallel via
\texttt{ProcessPoolExecutor}, with per-episode seeds derived deterministically
from a base seed. The same base seed always yields the same evaluation result,
making individual fitness calls reproducible.

\subsection{Tournament Evaluation}

Comparing candidate policies is formalized as a \texttt{Tournament}: a
cross-product of participants and scenarios, evaluated for $n_\text{ep}$
episodes each, aggregated by one of four voting rules.

\textbf{Argmax}: participant with the highest total score across all episodes.
\textbf{Majority voting}: pairwise---the participant winning more than half the
episodes in each pair advances; most pairwise wins overall.
\textbf{Borda count}: each participant receives points equal to the number of
opponents beaten per episode; totals summed.
\textbf{Copeland}: pairwise win/loss matrix; participant with most net wins.

HEAS reports the \emph{agreement matrix} (fraction of (scenario, repeat) pairs
where each pair of voting rules selects the same winner), the \emph{sample
complexity curve} $P(\text{correct})$ as a function of $n_\text{ep}$, and a
Kendall's~$\tau$~\cite{kendall1938new} vs.\ noise $\sigma$ stability curve.

\subsection{Statistical Infrastructure}

\texttt{heas/utils/stats.py} provides: \texttt{summarize\_runs()} (mean, std,
median, BCa bootstrap 95\% CI~\cite{efron1994bootstrap}),
\texttt{wilcoxon\_test()}, \texttt{cohens\_d()}, and
\texttt{kendall\_tau()} with bootstrap CI.
\texttt{heas/utils/pareto.py} provides \texttt{hypervolume()}~\cite{zitzler1998hypervolume}
and \texttt{auto\_reference\_point()}. All utilities accept \texttt{list[float]}
and return structured dicts, requiring no configuration.

% ============================================================
\section{CASE STUDIES}

We demonstrate HEAS's composability on two case studies that differ
structurally: an ecological population model with stochastic spatial dynamics
and a multi-firm economic regulation model with hierarchical governance.
Neither model is proposed as a validated scientific account of its
domain---both serve as \emph{composition vehicles} whose structural diversity
validates the generality of the Layer/Stream/Arena abstraction.

\subsection{Ecological Population Management}

The ecological model represents a fragmented landscape with prey and predator
populations interacting across climate-driven patch dynamics. It is implemented
as a five-stream Arena:

\begin{lstlisting}
EcoArena (200 steps/episode)
 +-- ClimateStream   -- temperature, shock_prob
 +-- LandscapeStream -- patch quality Q in [0,1]
 +-- PreyStream      -- logistic growth:
       dx = r*Q*x*(1-x/K) - loss(risk,x,y)
 +-- PredStream      -- dy = conv*x*y - mort*y
 +-- AggStream
       -> agg.mean_biomass (mean of x+y)
       -> agg.cv           (coefficient of variation)
       -> agg.extinct      (predator extinction flag)
\end{lstlisting}

The policy space is two-dimensional: \textbf{risk} $\in[0,1]$ and
\textbf{dispersal} $\in[0,1]$. The carrying capacity $K=1{,}000$ ensures
genuine predator-prey dynamics (breakeven prey $= \mathrm{mort}/(\mathrm{conv}
\cdot 0.01) = 750 < K$). At $K < 750$, the predator goes extinct in every
episode, making the extinction objective degenerate; detecting this required
observing HV~std$\,=0.000$ across 30 runs---a diagnostic only available via
the framework's reproducibility infrastructure.

NSGA-II minimizes $\{-\texttt{agg.mean\_biomass},\; \texttt{agg.cv}\}$
(maximize biomass, minimize variability). Tournament evaluation uses 8 scenarios
crossing fragmentation $\in\{0.2,\,0.6\}$ and
shock\_prob $\in\{0.05,\,0.15,\,0.20,\,0.30\}$.

\subsection{Enterprise Regulatory Design}

The enterprise model represents a regulated economy with firms, a sector
regulator, and a government. It is implemented as a four-layer Arena:

\begin{lstlisting}
EnterpriseArena
 +-- FirmLayer      -- production, compliance cost
 +-- RegulatorLayer -- audit, sector coordination
 +-- GovernmentLayer -- tax, subsidy allocation
 +-- WelfareLayer
       -> welfare.total, welfare.gini,
          welfare.compliance_rate
\end{lstlisting}

The four-dimensional policy space covers tax\_rate, audit\_intensity, subsidy,
and penalty\_rate. NSGA-II minimizes
$\{-\texttt{welfare.total},\;\texttt{welfare.gini}\}$. The 32-scenario
evaluation grid crosses governance regime, demand shock, audit frequency, firm
count, and cost structure (2 levels each).

\subsection{Structural Comparison}

Table~\ref{tab:structural} shows that the two case studies differ along every
structural axis. The metric contract (\texttt{metrics\_episode()} $\to$
\texttt{dict[str, float]}) generalizes from population dynamics to economic
welfare without any change in framework code.

\begin{table}[t]
\caption{Structural comparison of case studies.}
\label{tab:structural}
\centering\small
\begin{tabular}{lcc}
\toprule
Property          & Ecological     & Enterprise \\
\midrule
Hierarchy units   & 5 streams      & 4 layers \\
Policy genes      & 2              & 4 \\
Objectives        & biomass, CV    & welfare, Gini \\
Scenarios         & 8              & 32 \\
Episode length    & 200 steps      & 50 steps \\
Key dynamic       & stochastic     & deterministic \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{EVALUATION}

We evaluate HEAS along four dimensions: coupling code reduction versus Mesa
(the primary framework comparison), reproducibility of multi-run studies,
tournament reliability under voting rule variation and measurement noise, and
algorithm agnosticism across optimizer choices.

\subsection{Mesa Comparison: Coupling Code Reduction}
\label{sec:mesa-comparison}

We implement the ecological case study as a competent Mesa~3.3.1 model
(\texttt{experiments/mesa\_eco.py}) using all Mesa best practices:
\texttt{DataCollector} for metric logging, a separate episode runner function,
\texttt{ProcessPoolExecutor} for parallel evaluation, and individual fitness and
tournament scoring functions. \emph{This is not a strawman---it is the
implementation a Mesa-proficient researcher would produce.} We count non-blank,
non-comment lines in each component, separating model logic (identical between
implementations) from coupling code (lines that exist solely to wire simulation
output to a downstream consumer).

\begin{table}[t]
\caption{Coupling code LOC: Mesa~3.3.1 vs.\ HEAS.
  Model logic lines are excluded as they are identical in both.}
\label{tab:loc}
\centering\small
\begin{tabular}{lrrr}
\toprule
Task & Mesa & HEAS & Saved \\
\midrule
Metric contract / DataCollector setup    & 15  &  0  & 15 \\
Episode metric extraction function       & 20  &  0  & 20 \\
Multi-episode sequential runner          & 22  &  1  & 21 \\
Parallel episode runner                  & 20  &  0  & 20 \\
EA fitness function glue                 & 14  &  3  & 11 \\
Tournament scoring function              & 18  &  0  & 18 \\
Per-run seed management (30 runs)        & 12  &  0  & 12 \\
Bootstrap CI computation                 & 35  &  0  & 35 \\
Adding a second objective                &  4  &  1  &  3 \\
\midrule
\textbf{Total}                           & \textbf{160} & \textbf{5} & \textbf{155 (97\%)} \\
\bottomrule
\end{tabular}
\end{table}

The HEAS column's five lines correspond to one call to \texttt{run\_many()} in
the objective function body and four lines reading metric keys from the returned
dict. All other functionality---parallelism, seeding, Pareto tracking, tournament
voting, bootstrap CI---is provided by the framework at zero additional LOC.

\textbf{Objective extension cost} (Table~\ref{tab:extension}): adding a second
objective (biomass coefficient of variation) requires editing three files and
\textasciitilde{}6 lines in Mesa, with no framework mechanism to verify
consistency. HEAS requires one file and one line, with a structural consistency
guarantee: all consumers read the same \texttt{dict} key.

\begin{table}[t]
\caption{Cost of adding a second optimization objective.}
\label{tab:extension}
\centering\small
\begin{tabular}{lcc}
\toprule
                       & Mesa & HEAS \\
\midrule
Files edited           & 3    & 1    \\
Lines added            & 6    & 1    \\
Consistency guarantee  & None & Structural \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Metric divergence risk}: in Mesa, a developer who changes the EA
objective from \texttt{final\_prey} to \texttt{mean\_biomass} but does not update
the tournament scorer will silently obtain results where the optimization target
and evaluation criterion differ. In HEAS, this is structurally impossible: both
the EA and the tournament read \texttt{ep\_metrics["agg.mean\_biomass"]} from the
same \texttt{metrics\_episode()} return value.

\textbf{Parallelism}: \texttt{n\_jobs} to \texttt{run\_many()} is a zero-LOC
change in HEAS; Mesa requires \textasciitilde{}20 lines of
\texttt{ProcessPoolExecutor} boilerplate per project. For the lightweight ODE
simulation used here ($\approx$0.004\,s/episode), process startup overhead
($\approx$9\,s) dominates and \texttt{n\_jobs=4} provides no runtime speedup at
small episode counts. For computationally heavier agents ($>$0.1\,s/episode),
\texttt{n\_jobs} delivers linear speedup. The LOC ergonomics benefit is
unconditional.

\subsection{Reproducibility: 30-Run Statistical Studies}

We run 30 independent NSGA-II experiments per case study using framework-managed
per-run seeding and report hypervolume (HV) with 95\% BCa bootstrap CIs via
\texttt{summarize\_runs()}. Table~\ref{tab:hv30} shows the results.

\begin{table}[t]
\caption{Hypervolume statistics across 30 independent NSGA-II runs.
  Reference point: auto with 10\% margin.
  95\% BCa bootstrap CI.}
\label{tab:hv30}
\centering\small
\begin{tabular}{lcccc}
\toprule
Study       & HV mean  & HV std & 95\% CI & $n$ \\
\midrule
Ecological  & 7.665    & 3.518  & [6.424,\,8.914]       & 30 \\
Enterprise  & 4317.5   & 19.4   & [4311.2,\,4326.0]     & 30 \\
\bottomrule
\end{tabular}
\end{table}

The ecological HV distribution is bimodal: runs settle near local optima
(HV\,$\approx$\,4--6.5) or reach the global Pareto front
(HV\,$\approx$\,11.7--11.8). The CI width of 2.49 units correctly captures
this spread. The enterprise HV variance is narrow ($\pm$19.4) because the
enterprise landscape is more unimodal.

Crucially, the K$=1{,}000$ parameter selection was informed by this
reproducibility infrastructure: at $K=120$ (the original parameterization),
the predator went extinct in every episode, reducing both objectives to
degenerate constants and producing HV\,std\,$=0.000$ across all 30 runs. The
degeneracy was detected by observing the collapsed CI; a single-run Mesa study
would have reported one number without revealing the degenerate distribution.

\subsection{Tournament Validation}
\label{sec:tournament}

\textbf{Voting rule agreement} (Table~\ref{tab:voting}): we run 30 repeats of
the 8-scenario tournament with 100 episodes per scenario. Participants are:
champion (risk\,$=0.003$, dispersal\,$=0.959$, evolved by NSGA-II),
reference (risk\,$=0.55$, dispersal\,$=0.35$, paper baseline), and contrarian
(risk\,$=0.85$, dispersal\,$=0.15$, high-risk/low-dispersal). The champion's
mean biomass ($\approx$940) exceeds the reference ($\approx$785) by a margin
of $\approx$155 units. All four voting rules agree in 100\% of
(scenario,\,repeat) pairs.

\begin{table}[t]
\caption{Voting rule agreement matrix
  (fraction of trials, $n=30{\times}8=240$).}
\label{tab:voting}
\centering\small
\begin{tabular}{lcccc}
\toprule
          & Argmax & Majority & Borda & Copeland \\
\midrule
Argmax    & 1.000  & 1.000    & 1.000 & 1.000 \\
Majority  & ---    & 1.000    & 1.000 & 1.000 \\
Borda     & ---    & ---      & 1.000 & 1.000 \\
Copeland  & ---    & ---      & ---   & 1.000 \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Noise stability} (Table~\ref{tab:noise}): we inject Gaussian noise
$\mathcal{N}(0,\sigma^2)$ to episode scores and measure Kendall's~$\tau$
between the clean and noisy ranking. Rankings are stable ($\tau\!=\!1.0$) for
noise up to 0.65\% of the inter-policy margin, and degrade gracefully beyond.
Realistic simulation measurement variance is far below 10\% of typical
inter-policy margins, placing well-designed tournaments in the
$\tau\!>\!0.94$ regime.

\begin{table}[t]
\caption{Ranking stability under score perturbation
  ($n\!=\!30$ repeats, 50\,eps/scenario).
  $\sigma$/margin: noise relative to inter-policy score gap ($\approx$155).}
\label{tab:noise}
\centering\small
\begin{tabular}{rrcr}
\toprule
$\sigma$ & $\sigma$/margin & Mean~$\tau$ & 95\% CI \\
\midrule
0   &  0.0\%  & 1.000 & [1.000,\,1.000] \\
1   &  0.65\% & 1.000 & [1.000,\,1.000] \\
10  &  6.5\%  & 0.944 & [0.922,\,0.967] \\
50  & 32\%    & 0.744 & [0.700,\,0.789] \\
100 & 65\%    & 0.619 & [0.567,\,0.675] \\
200 & 130\%   & 0.508 & [0.453,\,0.567] \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Algorithm Agnosticism}
\label{sec:ablation}

To validate HEAS's optimizer-agnosticism claim, we run three strategies on the
ecological policy space (Table~\ref{tab:ablation}). Simple hillclimbing
outperforms NSGA-II on this 2-gene landscape: the Pareto front is effectively
one-dimensional (dominated by the dispersal gene), so single-objective search
concentrates population on the biomass-maximizing direction while NSGA-II's
diversity maintenance distributes population across the CV trade-off. This
result reinforces rather than undermines the framework claim: HEAS allows the
researcher to select the algorithm appropriate to the landscape complexity
without any framework changes.

\begin{table}[t]
\caption{Algorithm ablation on ecological 2-gene landscape
  ($n\!=\!10$ runs, pop\,=\,20, ngen\,=\,10).}
\label{tab:ablation}
\centering\small
\begin{tabular}{lccc}
\toprule
Strategy          & HV mean & HV std & 95\% CI \\
\midrule
Simple (hillclimb)& 19.66   & 6.84   & [14.82,\,22.92] \\
NSGA-II           &  9.99   & 6.84   & [6.67,\,14.81]  \\
Random grid       &  7.61   & 0.33   & [7.42,\,7.81]   \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{CONCLUSION}

HEAS addresses the coupling code problem in ABM-based policy search through
four architectural contributions: the Layer/Stream/Arena composition model,
the uniform metric contract, native NSGA-II integration, and the Tournament
evaluation primitive. Direct comparison with Mesa~3.3.1 confirms 97\% coupling
code reduction (5 vs.\ 160 lines); adding a second objective requires one line
in one file in HEAS versus six lines across three files in Mesa, with HEAS
providing structural consistency guarantees that Mesa cannot. Two case studies
validated across 30 independent runs each confirm framework correctness and
reproducibility. Tournament evaluation with four voting rules shows 100\%
agreement under well-separated participants; $\tau\!>\!0.94$ at realistic noise
levels confirms graceful degradation.

\textbf{Limitations.} For lightweight ODE simulations ($<$0.01\,s/episode),
process startup overhead ($\approx$9\,s) dominates and \texttt{n\_jobs}
provides no runtime benefit; use \texttt{n\_jobs=1}. HEAS is designed for
ODE-style and tabular agent simulations, not spatial grid simulations; for
spatially-explicit ABMs with rich movement and neighborhood interaction, Mesa
remains the preferred choice.

\textbf{Future work.} The highest-value near-term extension is a
\texttt{MesaLayerAdapter} that wraps a Mesa \texttt{Model} as a HEAS
\texttt{Layer}, exposing \texttt{DataCollector} reporters as
\texttt{metrics\_episode()} keys. This would make HEAS's EA and tournament
primitives available to existing Mesa models without rewriting the simulation
code. Additional planned work includes a Ray/Dask backend for HPC-scale parallel
evaluation, automated scenario generation via Latin Hypercube Sampling, and
adaptive hypervolume reference point estimation.

HEAS is available as an open-source Python package at
\url{https://github.com/ryZhangHason/heas}; all experiments in this paper are
fully reproducible from the scripts in \texttt{experiments/}.

% ============================================================
\section*{ACKNOWLEDGMENTS}

% Add acknowledgments here before submission.
% Example: "This work was supported by [grant agency and number]."
[Acknowledgments to be added.]

% ============================================================
\bibliographystyle{wsc}   % use the WSC bibliography style from Authors Kit
\bibliography{references}

% ============================================================
% AUTHOR BIOGRAPHIES — MANDATORY FOR WSC SUBMISSION
% One paragraph per author, must include email address.
% Given in the same sequence as listed under \author{}.
% ============================================================

\begin{biography}{[First Author]}
[First Author] is [title] in the [Department] at [Institution]. His/Her research
interests include agent-based simulation, evolutionary computation, and
simulation methodology. He/She is a member of [society]. He/She can be contacted
at \texttt{[email@institution.edu]}.
\end{biography}

\begin{biography}{[Second Author]}
[Second Author] is [title] in the [Department] at [Institution]. His/Her research
interests include multi-objective optimization, simulation frameworks, and
computational social science. He/She can be contacted at
\texttt{[email@institution.edu]}.
\end{biography}

\end{document}
